---
title: "Figures for Paper"
author: "Chela James"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding:
      hide
    toc: true
    toc_depth: 3
---

# Charlotta's paper
## Figures for Paper
### DGE summary
 - differences between fetal and adult cells in general

```{r include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = TRUE, warning = FALSE, message = FALSE)
```

```{r libraries}
library(knitr)
library(ggplot2)
library(sva)
library(dplyr)
library(preprocessCore)
library(ggrepel)
library(Biobase)
library(tibble)
library(scater)
library(gghighlight)
library(annotables)
library(DESeq2)
library(fgsea)
library(DT)
library(stringr)
library(hypeR)
library(qusage)
library(annotables)
library(clusterProfiler)
```

```{r data_prep}
# sample_table <- read.delim('~/projects/charlotta/cb_em/22_02_19/charlotta/human_primary_sample_22_03_19/charlotta_simplified_table.tsv', stringsAsFactors = FALSE)
# dir.create('data')
# save(sample_table, file = "data/sample_table.RData")
data(sample_table)
# prepare data from cluster
# jun_25_counts <- read.table('legion/human_big_counts', header = TRUE, stringsAsFactors = FALSE)
# nov_20_counts <- read.table('legion/20_nov_18/big_counts', header = TRUE, stringsAsFactors = FALSE)
# nov_20_counts <- nov_20_counts[which(rownames(nov_20_counts) %in% rownames(jun_25_counts)),]
# jun_25_counts<-jun_25_counts[which(rownames(jun_25_counts) %in% rownames(nov_20_counts)),]
# big_counts <- cbind(jun_25_counts,nov_20_counts)
# # 
# big_counts <- dplyr::rename(big_counts,
#                                BM_P5_11       = C23_N728_S505,
#                                BM_P10_11      = B04_N704_S503,
#                                BM_P12_11      = B06_N706_S503,
#                                CS19_P5_11     = B08_N710_S503,
#                                CS19_P12_11    = B12_N715_S503,
#                                BM1_P5_12      = B14_N718_S503,
#                                BM1_P10_12     = B16_N720_S503,
#                                BM1_P12_12     = B18_N722_S503,
#                                BM2_P5_12      = B20_N724_S503,
#                                BM2_P10_12     = B22_N727_S503,
#                                BM2_P12_12     = F20_N724_S508,
#                                CB_P5_12       = E23_N728_S507,
#                                CB_P10_12      = D04_N704_S506,
#                                CB_P12_12      = D06_N706_S506,
#                                CS22_P5_12     = D08_N710_S506,
#                                CS22_P10_12    = D10_N712_S506,
#                                CS22_P12_12    = D12_N715_S506,
#                                CS20_P5_12     = D14_N718_S506,
#                                CS20_P10_12    = D16_N720_S506,
#                                CS20_P12_12    = D18_N722_S506,
#                                FL13w_P5_13    = D20_N724_S506,
#                                FL13w_P10_13   = D22_N727_S506,
#                                FL13w_P12_13   = F22_N727_S508,
#                                CS23_P5_13     = G23_N728_S510,
#                                CS23_P10_13    = F04_N704_S508,
#                                CS23_P12_13    = F06_N706_S508,
#                                CS22_P5_19     = F08_N710_S508,
#                                CS22_P10_19    = F10_N712_S508,
#                                CS22_P12_19    = F12_N715_S508,
#                                w13_FL_P5_19   = F14_N718_S508,
#                                w13_FL_P10_19  = F16_N720_S508,
#                                w13_FL_P12_19  = F18_N722_S508)
# big_counts <- big_counts[,match(sample_table$Name, colnames(big_counts))]
# #  
# # # remove zero expression genes 
# big_counts <- big_counts[rowSums(big_counts) > 0,]
# save(big_counts, file = 'data/big_counts.RData')

rownames(sample_table) <- sample_table$Name
sample_table$Name <- NULL
sample_table$SampleID <- NULL
sample_table$batch <- c(rep("nov_18", 27), rep("mar_19",32))
sample_table$stage <- c("w12","w12","w10","w10","w10","CS19",
                        "CS19","CS19","CB","CB","CB","CS16",
                        "CS16","CB","CB","CB","w13","w13",
                        "CS20","CS20","CS20","w8","w8","w8",
                        "BM","BM","BM","BM","BM","BM","CS19"
                        ,"CS19","BM","BM","BM","BM","BM","BM",
                        "CB","CB","CB","CS22","CS22","CS22","CS20"
                        ,"CS20","CS20","w13","w13","w13","CS23",
                        "CS23","CS23","CS22","CS22","CS22",
                        "w13","w13","w13")


sample_table$pools <- c("W12_13","W12_13","W8_10","W8_10","W8_10","CS19_23","CS19_23","CS19_23","CB","CB","CB","CS16","CS16","CB","CB","CB","W12_13","W12_13","CS19_23","CS19_23","CS19_23","W8_10","W8_10","W8_10","BM","BM","BM","BM","BM","BM","CS19_23","CS19_23","BM","BM","BM","BM","BM","BM","CB","CB","CB","CS19_23","CS19_23","CS19_23","CS19_23","CS19_23","CS19_23","W12_13","W12_13","W12_13","CS19_23","CS19_23","CS19_23","CS19_23","CS19_23","CS19_23","W12_13","W12_13","W12_13")
sample_table$age <- case_when(str_detect(sample_table$pools,"W|CS") ~ "fetal", TRUE ~ "adult"  )

# remove CB and CS
sample_table <- sample_table[!str_detect(sample_table$pools,"CB|CS16"),]

# set diff groups to be adult and fetal
sample_table$batch <- str_replace(sample_table$batch, "mar", "jun")
datatable(sample_table,extensions = 'Buttons', options = list (dom = 'Bfrtip', buttons = c('copy', 'csv', 'excel'))) %>% 
  formatStyle("age", backgroundColor = styleEqual(c("fetal","adult"), c("#EAD3BF","#AA9486"))) %>%
  formatStyle("pools", backgroundColor = styleEqual(c("W12_13","CS19_23","CB","W8_10","BM","CS16"), c("#9986A5","#79402E", "#CCBA72","#F3DF6C","#D9D0D3","#8D8680")))
```

## Differential gene expression all adult v all fetal
### DGE performed after removing outliers

```{r}
# data(big_counts)
# sample_table2 <- sample_table[!rownames(sample_table) %in%
#                c("CB2_E4_8W_P12","CB2_E6_BM_P12","CB2_B2_CS19_P12",
#                                 "CB2_A7_10w_P12","CB2_D9_CS20_P12"),]
# 
# big_counts <- big_counts[, which(colnames(big_counts) %in% rownames(sample_table2))]
# dds <- DESeqDataSetFromMatrix(big_counts, colData = sample_table2, design = ~age)
# deseq <- DESeq(dds)
# save(deseq, file = 'data/deseq1.RData')
# data(deseq1)
# res <- results(deseq)
## clean up res
# res <- data.frame(res[complete.cases(res),])
# save(res, file = 'data/res1.RData')
## add entrez and gene symbols
data(res1)


my_colnames <- colnames(res)
res <- left_join(rownames_to_column(res,"ensgene"),grch38,by = "ensgene") %>%
  dplyr::select(ensgene,symbol,entrez,my_colnames)

# save for shiny app
# save(res,file="shiny/data/res.RData")
datatable(res[order(res$padj),], extensions = 'Buttons', options = list (dom = 'Bfrtip', buttons = c('copy', 'csv', 'excel')))
```

### Volcano plot 
#### All adult v all fetal
 
```{r}
res <- res %>% dplyr::mutate(sig = case_when( padj > 0.05 ~  1, padj < 0.05 ~ 0))
res$sig <- as.factor(res$sig)
ggplot(res,aes(log2FoldChange,-log10(padj),colour = sig)) + geom_point() +
  ylab(expression(paste("-", log[10], "(padj)", sep=""))) + scale_colour_manual(values = c("red","black")) + 
    xlab(expression(paste( log[2], "(fold change)", sep=""))) + theme_bw() + theme(legend.position = "none")
```

### GO 
- combine all changed genes between fetal and adult

```{r}
## genelist prepared according to these instructions:
## https://github.com/YuLab-SMU/DOSE/wiki/how-to-prepare-your-own-geneList
# 
HALLMARK <- msigdb_download_one(species="Homo sapiens", category="H")
geneList <- res$log2FoldChange
names(geneList) <- res$ensgene
geneList <- sort(geneList, decreasing = TRUE)
geneList <- geneList[!is.na(names(geneList))]
gene <- names(geneList)[abs(geneList) > 2]
# #
# library(org.Hs.eg.db)
# ego <- enrichGO(gene          = gene,
#                 universe      = names(geneList),
#                 OrgDb         = org.Hs.eg.db,
#                 keyType = "ENSEMBL",
#                 ont           = "BP",
#                 pAdjustMethod = "BH",
#                 pvalueCutoff  = 0.01,
#                 qvalueCutoff  = 0.05,
#         readable      = TRUE)
# # #
#save(ego, file = 'data/my_ego.RData')
data(my_ego)
barplot(ego, showCategory=20)
```

## GSEA using ProB
### Laurenti Muschen using ProB only

```{r}
# data(big_counts)
# sample_table2 <- sample_table[!rownames(sample_table) %in%
#                c("CB2_E4_8W_P12","CB2_E6_BM_P12","CB2_B2_CS19_P12",
#                                 "CB2_A7_10w_P12","CB2_D9_CS20_P12"),]
# 
# sample_table2 <- sample_table %>% tibble::rownames_to_column("my_rownames") %>% dplyr::filter(Phenotype == "CD34+CD19+" & pools == "BM" | Phenotype == "CD34+CD19+"  & pools == "CS19_23")
# 
# rownames(sample_table2) <- sample_table2$my_rownames
# big_counts <- big_counts[, which(colnames(big_counts) %in% rownames(sample_table2))]
# 
# dds <- DESeqDataSetFromMatrix(big_counts, colData = sample_table2, design = ~age)
# 
# 
# deseq3 <- DESeq(dds)
# save(deseq3, file = 'data/deseq3.RData')
data(deseq3)
res <- results(deseq3)
geneList <- res$log2FoldChange
names(geneList) <- rownames(res)
geneList <- sort(geneList, decreasing = TRUE)
geneList <- geneList[!is.na(names(geneList))]
laurenti <- qusage::read.gmt("Laurenti_Muschen.ensembl.gmt")
hyp_obj <- hypeR(geneList, laurenti, test="kstest", fdr_cutoff=0.01, do_plots = TRUE)
hyp_dots(hyp_obj )
```

### Hallmark with ProB yeilds nothing

```{r, eval = FALSE }
hallmark_hyp_obj <- hypeR(geneList, HALLMARK, test = "kstest", fdr_cutoff = 0.01, do_plots = TRUE)
hyp_dots(hallmark_hyp_obj)
```


### CMP

```{r}
hyp_obj$plots[[2]]
```


```{r}
hyp_obj$plots[[1]]
```


## Using IL7R only fetal v adult
### NO assotiation found with Laurenti gene lists.

```{r}
# data(big_counts)
# sample_table2 <- sample_table[!rownames(sample_table) %in%
#                c("CB2_E4_8W_P12","CB2_E6_BM_P12","CB2_B2_CS19_P12",
#                                 "CB2_A7_10w_P12","CB2_D9_CS20_P12"),]
# 
# sample_table2 <- sample_table2 %>% tibble::rownames_to_column("my_rownames") %>% dplyr::filter(Phenotype == "IL7R+Kit+" & pools == "BM" | Phenotype == "IL7R+Kit+"  & pools == "CS19_23")
# 
# rownames(sample_table2) <- sample_table2$my_rownames
# big_counts <- big_counts[, which(colnames(big_counts) %in% rownames(sample_table2))]
# 
# dds <- DESeqDataSetFromMatrix(big_counts, colData = sample_table2, design = ~age)
# 
# 
# deseq4 <- DESeq(dds)
# save(deseq4, file = 'data/deseq4.RData')
data(deseq4)
res <- results(deseq4)
geneList <- res$log2FoldChange
names(geneList) <- rownames(res)
geneList <- sort(geneList, decreasing = TRUE)
geneList <- geneList[!is.na(names(geneList))]
laurenti <- qusage::read.gmt("Laurenti_Muschen.ensembl.gmt")
# hyp_obj <- hypeR(geneList, laurenti, test="kstest", fdr_cutoff=0.01, do_plots = TRUE)
# hyp_dots(hyp_obj )
```

### Fetal v Adult IL7R only there is an enrichment in Hallmark genelists

```{r}
library(annotables)
my_genelist <- left_join(data.frame("ensgene" = names(geneList), "rank" = geneList), grch38, by = "ensgene" ) %>% dplyr::select(symbol, rank)
geneList = my_genelist$rank
names(geneList) <- my_genelist$symbol

hyp_obj <- hypeR(geneList, HALLMARK, test="kstest", fdr_cutoff=0.01, do_plots = TRUE)
hyp_dots(hyp_obj, show_plots=FALSE, return_plots=TRUE)
```


### G2M

```{r}
hyp_obj$plots[[1]]
```

### E2F Targets

```{r}
hyp_obj$plots[[2]]
```

## Epithelial to mesenchymal transition

```{r}
hyp_obj$plots[[3]]
```

### KRAS signaling DN

```{r}
hyp_obj$plots[[4]]
```

### hallmark myc targets

```{r}
hyp_obj$plots[[5]]
```


## GSEA with Adult V fetal (in general)

```{r}
data(res1)
res <- data.frame(res)
my_colnames <- colnames(res)
res <- left_join(rownames_to_column(res,"ensgene"),grch38,by = "ensgene") %>%
  dplyr::select(ensgene,symbol,entrez,my_colnames)
geneList <- res$log2FoldChange
names(geneList) <- res$symbol
geneList <- sort(geneList, decreasing = TRUE)
geneList <- geneList[!is.na(names(geneList))]

msigdb_path <- msigdb_download_all(species="Homo sapiens")
BIOCARTA <- msigdb_fetch(msigdb_path, "C2.CP.BIOCARTA")
KEGG     <- msigdb_fetch(msigdb_path, "C2.CP.KEGG")
REACTOME <- msigdb_fetch(msigdb_path, "C2.CP.REACTOME")


hyp_obj <- hypeR(geneList, HALLMARK, test="kstest", fdr_cutoff=0.01)
hyp_dots(hyp_obj, show_plots=FALSE, return_plots=TRUE)

```

## KEGG

```{r}
hyp_obj <- hypeR(geneList, KEGG, test="kstest", fdr_cutoff=0.01)
hyp_dots(hyp_obj, show_plots=FALSE, return_plots=TRUE)
```

## Reactome

```{r}
hyp_obj <- hypeR(geneList, REACTOME, test="kstest", fdr_cutoff=0.01)
hyp_dots(hyp_obj, show_plots=FALSE, return_plots=TRUE)
```

### UMAP/tSNE plot or PCA plot

- only IL7R and proB from BM and all fetal stages 
- exclude CB, do with and without CS16

### tSNE

```{r}
set.seed(123)
# UMAP, PCA and TSNE are embedded in scater
data(big_counts)
sample_table2 <- sample_table[!rownames(sample_table) %in%
               c("CB2_E4_8W_P12","CB2_E6_BM_P12","CB2_B2_CS19_P12",
                                "CB2_A7_10w_P12","CB2_D9_CS20_P12"),]

big_counts <- big_counts[, which(colnames(big_counts) %in% rownames(sample_table2))]

sce <- SingleCellExperiment(list(counts = as.matrix(big_counts)),
                            colData = sample_table2)

sce <- normalize(sce)
my_tsne_plot <- plotTSNE(sce, colour_by = "Phenotype", shape_by = "pools")

my_data <- my_tsne_plot$data
my_data <- my_data %>% dplyr::mutate(sample_id = rownames(.))
ggplot(my_data,aes(X,Y,colour = colour_by, shape = shape_by)) + 
  geom_point(size = 3) +
  theme_bw() +
  xlab(label = "dimension 1") +
  ylab(label = "dimension 2") + 
  labs(colour = "phenotype", shape = "pools") 
```

### PCA 

```{r}
my_pca_plot <- plotPCA(sce, colour_by = "batch", shape_by = "pools")

my_data <- my_pca_plot$data %>% dplyr::mutate(sample_id = rownames(.))
ggplot(my_data,aes(X,Y,colour = colour_by, shape = shape_by)) + 
  geom_point(size = 3) +
  theme_bw() +
  xlab(label = "PCA 1") +
  ylab(label = "PCA 2") + 
  labs(colour = "batch", shape = "pools") 
```


### PCA 3 dims

```{r}
plotPCA(sce, colour_by = "batch", shape_by = "pools", ncomponents = 3)
```

### UMAP

```{r}
my_umap_plot <- plotUMAP(sce, colour_by = "Phenotype", shape_by = "pools")
my_data <- my_umap_plot$data %>% dplyr::mutate(sample_id = rownames(.))
ggplot(my_data,aes(X,Y,colour = colour_by, shape = shape_by)) + 
  geom_point(size = 3) +
  theme_bw() +
  xlab(label = "UMAP 1") +
  ylab(label = "UMAP 2") + 
  labs(colour = "phenotype", shape = "pools") 
```

## Take out CD34+CD38-CD45RA-

### tSNE

```{r}
set.seed(123)
# UMAP, PCA and TSNE are embedded in scater
sce <- SingleCellExperiment(list(counts = as.matrix(big_counts)),
                            colData = sample_table2)

sce <- sce[,!str_detect(sce$Phenotype, "CD45RA")]

sce <- normalize(sce)
my_tsne_plot <- plotTSNE(sce, colour_by = "Phenotype", shape_by = "pools")

ggplot(my_tsne_plot$data,aes(x=my_tsne_plot$data$X,y=my_tsne_plot$data$Y,colour = colour_by, shape = shape_by)) + 
  geom_point(size = 3) +
  theme_bw() +
  xlab(label = "dimension 1") +
  ylab(label = "dimension 2") + labs(colour = "phenotype", shape = "pools")
```

### PCA

```{r}
my_pca_plot <- plotPCA(sce, colour_by = "Phenotype", shape_by = "pools")
ggplot(my_pca_plot$data,aes(x=my_pca_plot$data$X,y=my_pca_plot$data$Y,colour = colour_by, shape = shape_by)) + 
  geom_point(size = 3) +
  theme_bw() +
  xlab(label = "PCA 1") +
  ylab(label = "PCA 2") + labs(colour = "phenotype", shape = "pools")
```

### UMAP

```{r}
my_umap_plot <- plotUMAP(sce, colour_by = "Phenotype", shape_by = "pools")
ggplot(my_umap_plot$data,aes(x=my_umap_plot$data$X,y=my_umap_plot$data$Y,colour = colour_by, shape = shape_by)) + 
  geom_point(size = 3) +
  theme_bw() +
  xlab(label = "UMAP 1") +
  ylab(label = "UMAP 2") + labs(colour = "phenotype", shape = "pools")
```

## IL7R only

### tSNE

```{r}
set.seed(123)
# UMAP, PCA and TSNE are embedded in scater
sce <- SingleCellExperiment(list(counts = as.matrix(big_counts)),
                            colData = sample_table2)

sce <- sce[,str_detect(sce$Phenotype, "IL7R")]

sce <- normalize(sce)
my_tsne_plot <- plotTSNE(sce, colour_by = "pools", shape_by = "pools")

ggplot(my_tsne_plot$data,aes(x=my_tsne_plot$data$X,y=my_tsne_plot$data$Y,colour = colour_by, shape = shape_by)) + 
  geom_point(size = 3) +
  theme_bw() +
  xlab(label = "dimension 1") +
  ylab(label = "dimension 2") + labs(colour = "pools", shape = "pools")
```

### PCA

```{r}
my_pca_plot <- plotPCA(sce, colour_by = "pools", shape_by = "pools")
ggplot(my_pca_plot$data,aes(x=my_pca_plot$data$X,y=my_pca_plot$data$Y,colour = colour_by, shape = shape_by)) + 
  geom_point(size = 3) +
  theme_bw() +
  xlab(label = "PCA 1") +
  ylab(label = "PCA 2") + labs(colour = "pools", shape = "pools")
```

### DGE performed after HSCs taken out

```{r}
# data(big_counts)
# sample_table2 <- sample_table[!rownames(sample_table) %in%
#                c("CB2_E4_8W_P12","CB2_E6_BM_P12","CB2_B2_CS19_P12",
#                                 "CB2_A7_10w_P12","CB2_D9_CS20_P12"),]
# 
# big_counts <- big_counts[, which(colnames(big_counts) %in% rownames(sample_table))]
# dds <- DESeqDataSetFromMatrix(big_counts, colData = sample_table, design = ~age)
# dds <- dds[,dds$Phenotype != "CD34+CD38-CD45RA-"]
# deseq2 <- DESeq(dds)
# save(deseq2, file = 'data/deseq2.RData')
data(deseq2)

res <- results(deseq2)
## clean up res
res <- data.frame(res[complete.cases(res),])

## add entrez and gene symbols
my_colnames <- colnames(res)
res <- left_join(rownames_to_column(res,"ensgene"),grch38,by = "ensgene") %>%
  dplyr::select(ensgene,symbol,entrez,my_colnames)

datatable(res[order(res$padj),], extensions = 'Buttons', options = list (dom = 'Bfrtip', buttons = c('copy', 'csv', 'excel')))
```

### Volcano plot 
 IL7R prog and proB from adult BM vs w12-13, w8-10 and CS samples 
 - exclude CS16 for now
 
```{r}

res <- res %>% dplyr::mutate(sig = case_when( padj > 0.05 ~  1, padj < 0.05 ~ 0))
res$sig <- as.factor(res$sig)
ggplot(res,aes(log2FoldChange,-log10(padj),colour = sig)) + geom_point() +
  ylab(expression(paste("-", log[10], "(padj)", sep=""))) + scale_colour_manual(values = c("red","black")) + 
    xlab(expression(paste( log[2], "(fold change)", sep=""))) + theme_bw() + theme(legend.position = "none")

```

### GO 
- combine all changed genes between fetal and adult

```{r}
library(clusterProfiler)

## genelist prepared according to these instructions:
## https://github.com/YuLab-SMU/DOSE/wiki/how-to-prepare-your-own-geneList
# 
geneList <- res$log2FoldChange
names(geneList) <- res$ensgene
geneList <- sort(geneList, decreasing = TRUE)
geneList <- geneList[!is.na(names(geneList))]
gene <- names(geneList)[abs(geneList) > 2]
# #
library(org.Hs.eg.db)
ego <- enrichGO(gene          = gene,
                universe      = names(geneList),
                OrgDb         = org.Hs.eg.db,
                keyType = "ENSEMBL",
                ont           = "BP",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.01,
                qvalueCutoff  = 0.05,
        readable      = TRUE)
# # #
barplot(ego, showCategory=20)
```

## tryHyper

```{r}
library(hypeR)
library(annotables)
HALLMARK <- msigdb_download_one(species="Homo sapiens", category="H")
msigdb_path <- msigdb_download_all(species="Homo sapiens")
BIOCARTA <- msigdb_fetch(msigdb_path, "C2.CP.BIOCARTA")
KEGG     <- msigdb_fetch(msigdb_path, "C2.CP.KEGG")
REACTOME <- msigdb_fetch(msigdb_path, "C2.CP.REACTOME")

my_genelist <- left_join(data.frame("ensgene" = names(geneList), "rank" = geneList), grch38, by = "ensgene" ) %>% dplyr::select(symbol, rank)
geneList = my_genelist$rank
names(geneList) <- my_genelist$symbol

hyp_obj <- hypeR(geneList, HALLMARK, test="kstest", fdr_cutoff=0.01)
hyp_dots(hyp_obj, show_plots=FALSE, return_plots=TRUE)

```

## KEGG

```{r}
hyp_obj <- hypeR(geneList, KEGG, test="kstest", fdr_cutoff=0.01)
hyp_dots(hyp_obj, show_plots=FALSE, return_plots=TRUE)
```

## Reactome

```{r}
hyp_obj <- hypeR(geneList, REACTOME, test="kstest", fdr_cutoff=0.01)
hyp_dots(hyp_obj, show_plots=FALSE, return_plots=TRUE)
```


## FPKM looking at previous data

```{r, eval = FALSE}
fpkms1 <- read.table('legion/human_fpkm.tsv', header = TRUE, stringsAsFactors = FALSE )
fpkms2 <- read.table('legion/20_nov_18/temp2', header = TRUE, stringsAsFactors = FALSE )
```

